---
title: "Quick Processing"
output: 
  html_notebook:
    theme: flatly
    toc: true
    toc_float: true
---

# What is this?
E11.5 (or E11?) mouse RPE

Run on 10x v3 platform

This is a trial run to determine whether Elan can process mouse tissue at this very early time point.

```{r, fig.width=15, fig.height=15}
library(tidyverse)
library(Seurat)


seurat.data  <- Read10X(data.dir = '~/data/sc_mouse_OFC_e11/aggr/outs/count/filtered_feature_bc_matrix')
seurat <- CreateSeuratObject(counts = seurat.data, project = "e11_mouse_eye_cup", min.cells = 3, min.features = 200)
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, pattern = "^mt-")

seurat@meta.data$replicate <- str_extract(colnames(seurat), '\\d')
seurat@meta.data$orig.ident <- paste0(seurat@meta.data$orig.ident, '__', seurat@meta.data$replicate)
Idents(seurat) <- seurat@meta.data$orig.ident
```

# Basic stats per replicate

Many many more cells in rep4.
```{r}
seurat@meta.data %>% group_by(orig.ident) %>% summarise(Count = n()) %>% 
  ggplot(aes(x=orig.ident,y=Count, label = Count)) + geom_bar(stat = 'identity') +
  geom_text(position =position_nudge(y=200)) +
  coord_flip() +
  cowplot::theme_cowplot()
```

# Feature stats
But .... everything looks fairly the same between all four replicates.... so let's just keep it all for now? There's a larger size range in rep 4 of number of unique genes (nFeature_RNA) captured for some reason.
```{r}
VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, group.by = 'orig.ident')
```
# Filtering
Remove cells with percent.mt > 10%

```{r, fig.width=6}
seurat <- subset(seurat, subset = nFeature_RNA > 200  & percent.mt < 10)

seurat@meta.data %>% group_by(orig.ident) %>% summarise(Count = n()) %>% 
  ggplot(aes(x=orig.ident,y=Count, label = Count)) + geom_bar(stat = 'identity') +
  geom_text(position =position_nudge(y=200)) +
  coord_flip() +
  cowplot::theme_cowplot()
```
# PCA
```{r}
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000)



seurat <- FindVariableFeatures(seurat, selection.method = "vst", nfeatures = 2000)

all.genes <- rownames(seurat)
seurat <- ScaleData(seurat, features = all.genes)



seurat <- RunPCA(seurat, features = VariableFeatures(object = seurat), npcs = 50)
print(seurat[["pca"]], dims = 1:10, nfeatures = 10)


DimPlot(seurat, reduction = "pca")
```
# Heatmaps by PC
```{r, fig.width=16, fig.height=16}
DimHeatmap(seurat, dims = 1:30, cells = 500, balanced = TRUE)
```
# Jackstraw plot

Goal is to roughly figure out how many Principal Components (from PCA) to use downstream. I picked 20.
```{r}
seurat <- JackStraw(seurat, num.replicate = 100, dims = 50)
seurat <- ScoreJackStraw(seurat, dims = 1:50)


JackStrawPlot(seurat, dims = 1:50)


ElbowPlot(seurat, ndims = 50)
```
# Clustering and get markers for each cluster
```{r}
seurat <- FindNeighbors(seurat, dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.5)
seurat <- RunUMAP(seurat, dims = 1:20)


DimPlot(seurat, reduction = "umap")


seurat.markers <- list()
for (i in seurat@meta.data$seurat_clusters %>% unique()){
  seurat.markers[[i]] <- FindMarkers(seurat, ident.1 = i, only.pos = TRUE, logfc.threshold = 0.3)
}

```

# Regress out G2M <-> S phase diff

As the previous analysis essentially had clusters by cell phase, which is of less interest
```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes


m.s.genes <-s.genes %>% str_to_title()
m.g2m.genes <- g2m.genes %>% str_to_title()
seurat <- CellCycleScoring(seurat, s.features = m.s.genes, g2m.features = m.g2m.genes)

DimPlot(seurat, group.by = 'Phase')


seurat$CC.Difference <- seurat$S.Score - seurat$G2M.Score
options(future.globals.maxSize = 8000 * 1024^2)
seuratCC <- ScaleData(seurat, vars.to.regress = "CC.Difference", features = rownames(seurat))
seuratCC <- RunPCA(seuratCC, features = VariableFeatures(object = seuratCC), npcs = 50)
DimPlot(seuratCC, reduction = "pca")

seuratCC <- FindNeighbors(seuratCC, dims = 1:20)
seuratCC <- FindClusters(seuratCC, resolution = 0.5)
seuratCC <- RunUMAP(seuratCC, dims = 1:20)


DimPlot(seuratCC, reduction = "umap")
DimPlot(seuratCC, reduction = "umap", group.by = 'Phase')

seurat.markersCC <- list()
for (i in seuratCC@meta.data$seurat_clusters %>% unique()){
  seurat.markersCC[[i]] <- FindMarkers(seuratCC, ident.1 = i, only.pos = TRUE, logfc.threshold = 0.3)
}

```

# Output files
```{r}
# 2023-10-30 changed the logFC cutoff to 0.3 for the diff testing
save(seurat.markers, seurat, seurat.markersCC, seuratCC, file = '../data/00_process_output.freeze_2023_10_30.Rdata')
```

# Session
```{r}
sessionInfo()
```
