---
title: "Quick Analysis"
output: 
  html_notebook:
    theme: flatly
    toc: true
    toc_float: true
---

# What is this?
E11.5 (or E11?) mouse RPE

Run on 10x v3 platform

This is a trial run to determine whether Elan can process mouse tissue at this very early time point.

# When
2022-05-10

# Process Data
```{bash}
Rscript -e 'library(rmarkdown); rmarkdown::render("00_process.Rmd", "html_document")'
```

# Load Data

Quick peek at Elan's favorite Zfp503 and how which cluster it is differentially expressed in
```{r}
library(tidyverse)
library(Seurat)
load('../data/00_process_output.Rdata')
seurat.markersCC <- seurat.markersCC %>% bind_rows(.id = 'cluster') %>% as_tibble(rownames = 'Gene') %>% mutate(Gene = gsub('\\.\\.\\.\\d+','',Gene))
# seurat.markersCC %>% 
#   group_by(cluster) %>% slice_max(avg_log2FC, n = 3, with_ties = FALSE) %>% 
#   DT::datatable()
seurat.markersCC %>% bind_rows() %>%  filter(Gene == 'Zfp503')
```

# GO Terms
For each cluster

To help roughly figure out which clusters as which cell types
```{r, fig.height=7, fig.width=4}
library(clusterProfiler)
library(enrichplot)
eg_universe = bitr(seuratCC %>% row.names(), fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")

go_enrichCC <- list()
for (i in seurat.markersCC$cluster %>% unique()){
  print(i)
  diff_genes <- seurat.markersCC %>% filter(cluster  == i)
  eg_diff_genes <- bitr(diff_genes$Gene, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Mm.eg.db")
  eg_diff_genes <- diff_genes %>% left_join(., eg_diff_genes, by = c('Gene' = 'SYMBOL'))
  go_enrichCC[[i]] <- enrichGO(gene          = eg_diff_genes$ENTREZID,
                             universe      = eg_universe$ENTREZID,
                             OrgDb         = org.Mm.eg.db,
                             ont           = "all",
                             pAdjustMethod = "BH",
                             pvalueCutoff  = 0.01,
                             qvalueCutoff  = 0.05,
                             readable      = TRUE)
}



for (i in seurat.markersCC$cluster %>% unique() %>% sort()){
  print(dotplot(go_enrichCC[[i]], showCategory=30) + ggtitle(glue::glue("Dotplot for cluster {i}")))
}
 

```

# Hand label each cluster (guessing the cell type)
```{r}
ct_cluster_CC <- rbind(c(0, "RPC"),
  c(1, 'RPC'),
  c(2, 'RPE Precursor'),
  c(3, 'Fibroblast'),
  c(4, 'Fibroblast'),
  c(5, 'Optic Fissure'),
  c(6, 'Proliferating RPC'),
  c(7, 'Epithelial'),
  c(8, 'Red Blood Cell'),
  c(9, 'Corneal Progenitor'),
  c(10, 'Blood Vessel'),
  c(11, "PBMC"),
  c(12, "NK/T"),
  c(13, "Lens"
  )) %>% data.frame()
colnames(ct_cluster_CC) <- c('seurat_clusters','CellType')
seuratCC@meta.data$CellType <- seuratCC@meta.data %>% dplyr::select(seurat_clusters) %>% left_join(ct_cluster_CC) %>% pull(CellType)

# ct_cluster <- c(0, "RPC",
#   1, 'Fibroblasts',
#   2, 'RPC',
#   3, 'RPE Precursor',
#   4, 'Cornea - JCT - TM?',
#   5, 'Optic Fissure',
#   6, 'Proliferating RPC',
#   7, 'Epithelial',
#   8, 'Red Blood Cell',
#   9, '?',
#   10, 'Vessel',
#   11, "Macrophage/Microglia")

```

# UMAP
With my poorly researched cell type assignments
```{r, fig.width=5, fig.height=5}
DimPlot(seuratCC, reduction = "umap", group.by = 'CellType', label = TRUE, label.box = TRUE) + 
  scale_color_manual(values = pals::alphabet2() %>% unname()) +
  scale_fill_manual(values = pals::alphabet2() %>% unname() ) 
```

# Cell Phase 
```{r}
DimPlot(seuratCC, group.by = 'Phase')
```
# Markers
```{r, fig.width=8}
FeaturePlot(seuratCC, features = c("Pax2", "Ntn1", "Smoc2", "Vax1", 
                                 "Ttr", "Dct", "Rho", "Zfp503", "Mitf",
                                 "Pax6"))
```

# Top 10 Cluster 5 (Optic Fissure?) Diff Genes
```{r, fig.width=8}

top_genes <- seurat.markersCC %>%  
  left_join(seuratCC@meta.data %>% as_tibble() %>% 
              dplyr::select(cluster = seurat_clusters, CellType) %>% unique()) %>% 
  group_by(cluster, CellType) %>% top_n(30, avg_log2FC) %>% 
  filter(cluster == 5) %>% head(15) %>% pull(Gene)
FeaturePlot(seuratCC, features = top_genes)
```

# Zfp503 Like Genes

Zfp503 is overexpressed in both clusters 2 and 5

What other genes have this characteristic?
```{r, fig.width=8}

 seurat.markersCC %>% filter(cluster %in% c(2,5), avg_log2FC > 0.7, p_val_adj < 0.01) %>% 
  group_by(Gene) %>% 
  summarise(Count = n(), avg_log2FC = mean(avg_log2FC)) %>% 
  filter(Count == 2)

FeaturePlot(seuratCC, features = seurat.markersCC %>% filter(cluster %in% c(2,5), avg_log2FC > 0.7, p_val_adj < 0.01) %>% 
  group_by(Gene) %>% 
  summarise(Count = n(), avg_log2FC = mean(avg_log2FC)) %>% 
  filter(Count == 2) %>% 
    pull(Gene))

```

# Cluster diff expressed genes

Top 30 gene for each cluster (I was using this table along with the GO terms to guess each cell type for each cluster)
```{r}
seurat.markersCC %>%  
  left_join(seuratCC@meta.data %>% as_tibble() %>% 
              dplyr::select(cluster = seurat_clusters, CellType) %>% unique()) %>% 
  group_by(cluster, CellType) %>% top_n(30, avg_log2FC) %>% 
  relocate(CellType) %>% 
  arrange(cluster) %>% 
  DT::datatable()
```

# Quick comments

1. The experiment appears to have worked
2. You seem to have a Optic Fissure specific cluster?? Cluster 5. Several interesting genes in that cluster.

# Output Cluster and UMAP for Loupe
```{r}
write_csv(seuratCC@reductions$umap@cell.embeddings %>% as_tibble(rownames = 'Barcode'), file = 'umap.loupe.csv')
write_csv(seuratCC@meta.data %>% as_tibble(rownames = 'Barcode') %>% dplyr::select(Barcode, seurat_clusters), file = 'cluster.loupe.csv')
```

# Sesssion Info
```{r}
devtools::session_info()
```
