---
title: "Prototype Figures"
output: 
  html_notebook:
    theme: flatly
    toc: true
    toc_float: true
---


# Load Data

```{r}
library(tidyverse)
library(Seurat)
load('../data/00_process_output.Rdata')
seurat.markersCC <- seurat.markersCC %>% bind_rows(.id = 'cluster') %>% as_tibble(rownames = 'Gene') %>% mutate(Gene = gsub('\\.\\.\\.\\d+','',Gene))
```

# Hand label each cluster (guessing the cell type)
```{r}
ct_cluster_CC <- rbind(c(0, "RPC"),
                       c(1, 'RPC'),
                       c(2, 'RPE Precursor'),
                       c(3, 'Fibroblast'),
                       c(4, 'Fibroblast'),
                       c(5, 'Optic Fissure'),
                       c(6, 'Proliferating RPC'),
                       c(7, 'Epithelial'),
                       c(8, 'Red Blood Cell'),
                       c(9, 'Corneal Progenitor'),
                       c(10, 'Blood Vessel'),
                       c(11, "PBMC"),
                       c(12, "NK/T"),
                       c(13, "Lens"
                       )) %>% data.frame()
colnames(ct_cluster_CC) <- c('seurat_clusters','CellType')
seuratCC@meta.data$CellType <- seuratCC@meta.data %>% dplyr::select(seurat_clusters) %>% left_join(ct_cluster_CC) %>% pull(CellType)

# ct_cluster <- c(0, "RPC",
#   1, 'Fibroblasts',
#   2, 'RPC',
#   3, 'RPE Precursor',
#   4, 'Cornea - JCT - TM?',
#   5, 'Optic Fissure',
#   6, 'Proliferating RPC',
#   7, 'Epithelial',
#   8, 'Red Blood Cell',
#   9, '?',
#   10, 'Vessel',
#   11, "Macrophage/Microglia")

```

# UMAP
With my poorly researched cell type assignments
```{r, fig.width=10, fig.height=10}
DimPlot(seuratCC, reduction = "umap", group.by = 'CellType', label = TRUE, label.box = TRUE) + 
  scale_color_manual(values = pals::alphabet2() %>% unname()) +
  scale_fill_manual(values = pals::alphabet2() %>% unname() ) + theme(legend.position = "none")
```


```{r, fig.width=5, fig.height=5}
Seurat::FetchData(seuratCC, vars = c("UMAP_1","UMAP_2", "CellType")) %>%
  ggplot(aes(x=UMAP_1,
             y=UMAP_2, 
             color = CellType, 
             label = CellType)) +
  cowplot::theme_cowplot() +
  scale_color_manual(values = pals::glasbey() %>% unname()) +
  scale_fill_manual(values = pals::glasbey() %>% unname() ) + theme(legend.position = "none") +
  geom_point(size = 0, alpha = 0.8) +
  geom_text_repel(data = . %>% group_by(CellType) %>%
                    summarise(UMAP_1 = mean(UMAP_1),
                              UMAP_2 = mean(UMAP_2)),
                  color = 'gray10',
                  bg.color = 'gray90') +
  xlab("UMAP 1") + ylab("UMAP 2")
```

# Markers
```{r, fig.width=8}
FeaturePlot(seuratCC, features = c("Pax2", "Ntn1", "Smoc2", "Vax1", 
                                   "Ttr", "Dct", "Rho", "Zfp503", "Mitf",
                                   "Pax6"))
```

# Top 10 Cluster 5 (Optic Fissure?) Diff Genes
```{r, fig.width=8}

top_genes <- seurat.markersCC %>%  
  left_join(seuratCC@meta.data %>% as_tibble() %>% 
              dplyr::select(cluster = seurat_clusters, CellType) %>% unique()) %>% 
  group_by(cluster, CellType) %>% top_n(30, avg_log2FC) %>% 
  filter(cluster == 5) %>% head(15) %>% pull(Gene)
FeaturePlot(seuratCC, features = top_genes)
```

# Zfp503 Like Genes

Zfp503 is overexpressed in both clusters 2 and 5

What other genes have this characteristic?
```{r, fig.width=8}

seurat.markersCC %>% filter(cluster %in% c(2,5), avg_log2FC > 0.7, p_val_adj < 0.01) %>% 
  group_by(Gene) %>% 
  summarise(Count = n(), avg_log2FC = mean(avg_log2FC)) %>% 
  filter(Count == 2)

FeaturePlot(seuratCC, features = seurat.markersCC %>% filter(cluster %in% c(2,5), avg_log2FC > 0.7, p_val_adj < 0.01) %>% 
              group_by(Gene) %>% 
              summarise(Count = n(), avg_log2FC = mean(avg_log2FC)) %>% 
              filter(Count == 2) %>% 
              pull(Gene))

```

# Cluster diff expressed genes

Top 30 gene for each cluster (I was using this table along with the GO terms to guess each cell type for each cluster)
```{r}
seurat.markersCC %>%  
  left_join(seuratCC@meta.data %>% as_tibble() %>% 
              dplyr::select(cluster = seurat_clusters, CellType) %>% unique()) %>% 
  group_by(cluster, CellType) %>% top_n(30, avg_log2FC) %>% 
  relocate(CellType) %>% 
  arrange(cluster) %>% 
  DT::datatable()
```

# Quick comments

1. The experiment appears to have worked
2. You seem to have a Optic Fissure specific cluster?? Cluster 5. Several interesting genes in that cluster.

# Output Cluster and UMAP for Loupe
```{r}
write_csv(seuratCC@reductions$umap@cell.embeddings %>% as_tibble(rownames = 'Barcode'), file = 'umap.loupe.csv')
write_csv(seuratCC@meta.data %>% as_tibble(rownames = 'Barcode') %>% dplyr::select(Barcode, seurat_clusters), file = 'cluster.loupe.csv')
```

# Sesssion Info
```{r}
devtools::session_info()
```
